---
title: "UWPCE DataScience3 KaggleWNVP Project Report"
author: "Kaggle Team 3"
date: "2 June 2015"
output: pdf_document
geometry: margin=0.25in
fontsize: 12pt
mainfont: Verdana
monofont: Consolas
---

# Background

WNVP: "West Nile Virus Prediction"

[URL of Kaggle Competition](https://www.kaggle.com/c/predict-west-nile-virus).

Goal of Competition: "Predict West Nile virus in mosquitos across the city of Chicago"

Team Goals:

1. Experiment with modelling alternative on a real-world dataset.
2. Learn how to participate in a Kaggle competition.
3. Non-Goal: win the competition, or even score highly.

# Team Members

* Bethene Britt
* Andrew Ewing
* Gregory Hogue
* Patrick Leahy
* Michal Monselise
* Linghua Qiu
* Chris Ross
* Robert Russell
* Jim Stearns

# Data Preparation

## Obtain Original Datasets from Kaggle Website

Download the training, test, spray, and weather data from the Kaggle web site page for West Nile Virus Prediction.

Please see Appendix for R (and Python) code that:

* General Setup: Clears environment, sets working directory, loads libraries
* downloads the data from the Kaggle site and unzips them. 

```{r Setup, echo=FALSE, results='hide',message=FALSE}
# Clear the working environment of variables, data, functions
rm(list=ls())

# Set working directory for this Kaggle project. Default: pwd.
#kaggleProjHomeDir <- "."
kaggleProjHomeDir <- "/Users/jimstearns/GoogleDrive/Learning/Courses/UWPCE-DataScience/Course3_DataAtScale/KaggleProject/Report"
setwd(kaggleProjHomeDir)
getwd()

#install.packages("rPython")  # For download from web site with login/pwd.
library(rPython) # For calling python function to download file w/login+pwd
```

```{r GetDatasetsIntoPwd, echo=FALSE, results='hide',message=FALSE}
wnvpTrainFilename <- "train.csv"
wnvpTestFilename <- "test.csv"
wnvpWeatherFilename <- "weather.csv"
wnvpSprayFilename <- "spray.csv"
kaggleDatasets = c(
    wnvpTrainFilename, 
    wnvpTestFilename, 
    wnvpWeatherFilename, 
    wnvpSprayFilename)
dataSubDir= "Data"

# If download from Kaggle required, and user and pwd are empty (default),
# then user will be prompted for these two values.
kaggleUsername <- ""
kagglePassword <- ""

allKaggleFilesArePresent <- function() {
    filesAllFound <- TRUE
    for (file in kaggleDatasets) {
        if (!file.exists(paste0(dataSubDir, "/", file))) {
            print(paste("Error: could not find unzipped Kaggle file in PWD:", file))
            filesAllFound <- FALSE
        }
    }
    return(filesAllFound)
}

downloadMissingKaggleFiles <- function() {
    python.load("UrlFileDownloaderWithLogin.py")
    
    kaggleUsername = Sys.getenv("kaggleUsername")
    kagglePassword = Sys.getenv("kagglePassword")
    if (kaggleUsername == "" || kagglePassword == "") {
        print("Please assign kaggleUsername and kagglePassword environment variables.")
        print("Place in ~/.Renviron entries such as kaggleUsername='YourName'.")
    }
    stopifnot(!(kaggleUsername == ""))
    stopifnot(!(kagglePassword == ""))
    
    wnvpKaggleDataUrl <- 
        "https://www.kaggle.com/c/predict-west-nile-virus/download/"
    
    for (file in kaggleDatasets) {
        if (file.exists(file))
            next
        
        urlOfZip <- paste0(wnvpKaggleDataUrl, file, ".zip")
        print(urlOfZip)
        # Use a python method to download from URL with login and password.
        # Download to subdirectory Data and filename w/o the .zip suffix.
        python.call("Download", urlOfZip, 
                    kaggleUsername, kagglePassword , 
                    paste0(dataSubDir, "/", file, ".zip"))
    }
}

unzipDownloadedFiles <- function() {
    for (file in kaggleDatasets) {
        zippedFile <- paste0(dataSubDir, "/", file, ".zip")
        print(paste0("Unzip: ", zippedFile))
        if (file.exists(zippedFile)) {
            if (file.exists(file)) {
                print(sprintf("Warning: removing existing file %s\n", file))
                file.remove(file)
            }
            unzip(zippedFile, exdir=dataSubDir)
            print(sprintf("Unzipped: %s\n", zippedFile))
        }
    }
}

if (!allKaggleFilesArePresent()) {
    print(paste("Not all needed Kaggle datasets are present in PWD;",
                "attempting to download from Kaggle web site."))
    downloadMissingKaggleFiles()
    unzipDownloadedFiles()
}

```

```{r}
stopifnot(allKaggleFilesArePresent())
print("All unzipped Kaggle datasets found in PWD. Proceeding.")
```

# Appendix

## General Setup: Clear environment, set working directory, load libraries

```{r Setup,eval=FALSE}
```

## Dataset download and unpacking

This R and Python code downloads the WNVP datasets from Kaggle. 
Some setup is required:

* One's Kaggle username and password must be
defined as environment variables where R is running.
* Easiest way to set environment variable for R: Create (add to) ~/.Renviron file (kaggleUsername="XXXX" and kagglePassword="YYYY").

Alternatively, files can be downloaded manually.
  
```{r GetDatasetsIntoPwd,eval=FALSE}
``` 

File UrlFileDownloaderWithLogin.py:
```
__author__ = 'jimstearns'
""" Download a file at a URL at a web site that requires a user name and password.
"""

import logging
import os       # File utilities

# Python package "requests": "Python HTTP for Humans" by Kenneth Reitz. Current version: 2.7.0.
# Documented at http://docs.python-requests.org/en/latest/
# To install from the command line: "pip install requests"
# (On Mac, sudo may be required. Also pip2.7 instead of pip, depending on default Python version)

import requests # Http GET, POST

def Download(url, username, password, local_filename):
    # Login to web site such as Kaggle and retrieve the data. Use POST rather than GET as as to
    # send login info in body of HTTP request rather than in query string portion of URL.
    
    # Limitation: when used by Python version < 2.7.9, an "InsecureRequestWarning" is generated.
    # TODO: Fix. Details: https://urllib3.readthedocs.org/en/latest/security.html#insecureplatformwarning
    # Workaround: log warnings to file, not stdout.
    logging.captureWarnings(True)

    if (os.path.exists(local_filename)):
        os.remove(local_filename)

    # This won't get the file, but use the return value URL in a follow-on POST:
    r = requests.get(url)

    login_info = {'UserName': '{0}'.format(username), 'Password': '{0}'.format(password) }
    print(login_info)
    r = requests.post(r.url, data = login_info)
    print("POST (w/login info): {0}\n".format(r.status_code))

    # Write the data to a local file one chunk at a time.
    chunk_size = 512 * 1024 # Reads 512KB at a time into memory
    with open(local_filename, 'wb') as fd:
        for chunk in r.iter_content(chunk_size): # Reads 512KB at a time into memory
            if chunk: # filter out keep-alive new chunks
                fd.write(chunk)

    if (os.path.exists(local_filename)):
        return(True)
    else:
        return(False)
```